Hydras \& Co.

### 2023-05-23 doc/chapter-fol.tex
redaction of chapter 12 (in progress)

This chapter is devoted to the presentation to  data structures for representing first order formulas over a ranked alphabet, and the basic functions and predicates over these types.
Although all the following constructions come directly from~\cite{Goedel}, we introduced minor changes (mainly syntax) to
take into account recent changes in \coq(new constructions, tactics, notations, etc.).

@@

A \emph{language} is a structure composed of relation and function symbols, each symbol is given an \emph{arity} (number of arguments)\,\footnote{As suggested by Russel O'Connor in~\cite{OConnor05}, we consider two arity functions instead of a function defined on the sum type \texttt{Relations + Functions}}.

@@

\begin{remark}[Variables]
In O'Connor's formalization of first-order logic,  variables are \gallina terms of the form (\texttt{var $L$ $i$}), where $L$ is
 a language and $i$ a natural number. 
Although other choices may be considered : PHOAS, de Bruijn indices, etc,  we   respect this representation, in order not to break long proof scripts which use this representation (please look at \cite{OConnor05} for a related discussion).

@@

@@

@@

Let us define some \gallina terms over our toy language, respecively corresponding to the terms $a$,
$f(a)$, and $h(f(a),a)$. The fourth ``term'' is not well formed, since the arity of $h$ is not respected\,\footnote{Strictly speaking, it's not a (well typed) term!}.

\subsection{Formulas}

The  abstract syntax of first-order formulas is completed 
with a few abbreviations\,\footnote{Technically, these abbreviations take the form of \emph{global definitions}}. Note that we are studying \emph{classical} first-order logic, for instance the existential quantifier is derived from the universal one through negation.

Let us give a few examples of first-order formulas over $L$.

\label{fol:examplesf1f2f3}
\begin{enumerate}
\item $R\;a\;b$
\item $\forall v_0\; v_1, R\;v_0\;v_1 \leftrightarrow R\;v_1\;v_0$
\item $ \forall v_0, v_0=a \vee \exists\;v_1,\, v_0= f(v_1)$
\end{enumerate}

\label{sect:folFormExamples}

\section{Simplified notations for terms and formulas}
\label{sect:fol-notations}

 We use \coq's \texttt{Notation} features to print and parse terms and formulas  in a more readable form.

 \subsection{Parsing and printing terms}
 \label{sec:fol-term-notations}

Let us consider the term $t_2$ of Sect\ref{sect:folTermExamples}~. Its normal form is quite hard to read:

\inputsnippets{FolExamples/beforeAbbreviations}

We may use \coq's abbreviation feature  to print and parse $L$-terms with a more natural concrete syntax.

\inputsnippets{FolExamples/termAbbreviations, FolExamples/NowItsBetter}

\subsection{Parsing and printing first-order formulas}

In order to make the first order formulas more readable, we define  a  notation scope called \texttt{fol\_scope}, which contains 
notations for connectives and quantifiers. The \texttt{\%fol} delimiter 
allows the user to distinguish \texttt{FOL} connectives from their \coq equivalent.

@@

The following notations are used (when printing) when  some computation 
    expands a disjunction, conjuction, etc. 
    in terms of implication and negation.

@@

\subsubsection{Simplified notations for atomic formulas}
\label{sec:fol-atomic-notations}

In addition to \texttt{fol\_scope}, we can introduce a few abbreviations specific to our toy language.

The following interactions show how the formulas 
of~\ref{sect:folFormExamples} may be 
parsed or printed.

>>>>

This chapter is devoted to the presentation to  data structures for representing first order formulas over a ranked alphabet, and the basic functions and predicates over these types:

\begin{itemize}
\item Abstract syntax of terms and formulas over a ranked alphabet composed of function and relation symbols.
\item Induction principles over terms and formulas.
\item Definition and main properties of substitution of terms to variables.
\end{itemize}

Although all the following constructions come directly from~\cite{Goedel}, we introduced minor changes (mainly syntax) to
take into account recent changes in \coq(new constructions, tactics, notations, etc.).

@@

A \emph{language} is a structure composed of relation and function symbols, each symbol is given an \emph{arity} (number of arguments)\,\footnote{As suggested by Russel O'Connor in~\cite{OConnor05}, we consider two arity functions instead of a single function defined on the sum type \texttt{Relations + Functions}.}.

@@

\begin{remark}[Variables]
In O'Connor's formalization of first-order logic,  variables are 
just natural numbers, and the \emph{coercion} from
\texttt{nat} to \texttt{Term L} is the constructor (\texttt{@var $L$}).
Although other choices may be considered : PHOAS, de Bruijn indices, etc,  we still the data structures of~\cite{Goedel}, in order not to break long proof scripts which use this representation (please look at \cite{OConnor05} for a related discussion).

@@

Let us define some \gallina terms over our toy language, respecively corresponding to the terms $a$,
$f(a)$, and $h(f(a),a)$.

First, in order to make terms on $L$ more readable, we introduce a few abbreviations.

\inputsnippets{FolExamples/termAbbreviations}

\begin{remark}
The fourth ``term'' \texttt{t4} is not well formed, since the arity of $h$ is not respected\,\footnote{Strictly speaking, it's not a (well typed) term!}.
\end{remark}

\subsubsection{Other Languages}

\begin{todo}
Link to the chapter which presents \texttt{LNT} and \texttt{LNN}.
\end{todo}

\subsection{First-order formulas}

\begin{remark}
The type \texttt{Formula} (over $L$) has only five constructors,
respectively bound to term equalities $t_1=t_2$, atomic formulas of the form $(r\;t_1\;\dots\;t_n)$, implications $(F \arrow G)$, negations $({\sim}F)$, and universal quantifications $(\forall\,i, F)$.

In~\cite{Goedel}, no \emph{constructors} are associated with
disjunction, conjonction, logical equivalence and existential quantifier. These constructs are formalized through \emph{definitions}\footnote{Please keep in mind that we are considering a classical logic.}.

This convention allows the user to reduce to 5 (instead of 10) the number of cases in ''\texttt{match F with \dots}'' constructs. On the other hand, some computation may expand a connective like $\vee$ or
$\wedge$, or an existential quantification into a ``basic'' formula (see Sect.\vref{sect:fol-issue}).
\end{remark}

Let us give a few examples of first-order formulas over $L$.
\label{fol:examplesf1f2f3}
\begin{enumerate}
\item $R\;a\;b$
\item $\forall v_0\; v_1, R\;v_0\;v_1 \arrow R\;v_1\;v_0$
\item $ \forall v_0, v_0=a \vee \exists\;v_1,\, v_0= f(v_1)$
\item $(\forall v_1, v_0 = v_1) \vee \exists\,v_0\,v_1, v_0 \not= v_1$
\end{enumerate}

First, we define abbreviations for atomic formulas over $L$.

\label{sec:fol-atomic-notations}
\label{sect:folFormExamples}

% The following interactions show how the formulas 
% of~\ref{sect:folFormExamples} may be 
% parsed or printed.

% \noindent From~\href{../theories/html/hydras.MoreAck.FolExamples.html}{MoreAck.FolExamples}

% \vspace{6pt}

% \inputsnippets{FolExamples/toyNotationForm2}

@@

In the formula \texttt{F4} \vpageref[above]{sect:folFormExamples}, the  term \texttt{(var 0)} occurs freely and also inside the scope of a quantifier.

@@

\section{A notation scope for first-order terms and formulas}
\label{sect:fol-notations}

 We use \coq's \texttt{Notation} features to print and parse terms and formulas  in a more readable form.
To this purpose, we build \texttt{fol\_scope}, a  notation scope
where the main connectives and quantifiers get a syntax close to \coq's.  Additionnally, a term of the form \texttt{(var $i$)} is
just written \texttt{v\#$i$}.

The \texttt{\%fol} delimiter 
allows the user to distinguish \texttt{FOL} connectives from their \coq equivalent.
\emph{We discourage the reader from \emph{opening} \texttt{fol\_scope} and similar scopes : \texttt{nn\_scope}, \texttt{nt\_scope}, which would make disappear the \texttt{\%fol} suffix from the first-order formulas}.

@@

\subsection{The issue with derived constructions}
\label{sect:fol-issue}

The connectives and quantifiers $\vee$, $\wedge$, $\exists$, etc. may raise an issue when printing computed formulas.
For instance, a formula like $F \wedge G$ could be transformed into $\sim(\sim F \vee \sim G)$,  and even
into $\sim(\sim\sim F \arrow \sim B)$, which would 
cause serious problems of readability.

In such a case, we propose to print such a formula as $F \wedge' G$, to make it syntactically distinct but very similar to $F \wedge G$.

@@

The following examples show how the primed connectors 
and quantifiers behave with respect to convertibility and 
input/output.

\inputsnippets{FolExamples/toyNotationForm3}
